Towards topological type theory for decrypting transfinite methods in classical mathematics

### 2025-06-20 slides-types2025.lagda.tex
Notes slides

\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

% Workaround for the issue described at
% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
\newcommand{\fixedhref}[2]{\makebox[0pt][l]{\hspace*{\paperwidth}\href{#1}{#2}}\href{#1}{#2}}

@@

\title{Modal operators for a constructive account of well quasi-orders}

\author{Ingo Blechschmidt}
\date{November 30th, 2024}

%\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}

@@

\newunicodechar{∇}{\ensuremath{\nabla}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{∷}{\ensuremath{::}}
\newunicodechar{⧺}{\ensuremath{\!\!\!}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{ℕ}{\ensuremath{\NN}}
\newunicodechar{↠}{\ensuremath{\twoheadrightarrow}}
\newunicodechar{ʳ}{\ensuremath{^r}}

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{A case study in Hilbert's program}
  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
  \vspace*{-0.3em}
  \begin{itemize}
    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
  \end{itemize}
  \pause
  \vspace*{-1.3em}

@@

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
  the following inductively defined predicate on \hil{finite lists}.~\formalized{}
  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
  \text{well}\sinf$.)
  \[
    \small
    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
    \qquad
    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma ::^r x)}
  \]
  \vspace*{-1.3em}

  \only<8>{
    \begin{columns}[c]
      \begin{column}{0.01\textwidth}
	\includegraphics[height=2.4em]{question-mark}
      \end{column}
      \begin{column}{0.9\textwidth}
      Is there a procedure for reinterpreting \hil{classical proofs}
      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
      \end{column}
    \end{columns}
  }

  \only<9>{
    \vspace*{-0.0em}
    \hil{Central insight:} A quasi-order $X$ is well\ind iff $\hil{$\necessary$}\,\forall\alpha : \NN \to
    X\_ \exists i < j\_ \alpha\,i \leq \alpha\,j$.
  }
\end{frame}}

\begin{frame}{Missing functions in the type of all functions?}
  \begin{columns}[t]
    \begin{column}{0.57\textwidth}
      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
        \justifying
        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
        Let~$\alpha : \NN \to X$. Then there is an increasing
        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
      \end{block}

@@

    In set and type theory, we can also enlarge \hil{the mathematical universe}:
    \begin{itemize}
      \small
      \item Force a \hil{generic sequence} $\NN \to X$. \formalized
      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
            (even if $X$ is uncountable). \formalized
      \item Force a \hil{generic prime ideal} of a given ring. \formalized
    \end{itemize}
    \pause

    Central observations of the multiversal yoga:
    \begin{itemize}
      \small
      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
      is good. \formalized
      \item A relation is well-founded iff for the generic descending chain,
      $\bot$. \formalized
      \item A ring element is nilpotent iff it is contained in the generic
      prime ideal. \formalized
    \end{itemize}
    \pause

    The mystery of nongeometric sequents:
    \begin{itemize}
    \small
      \item The generic ring is a field.
      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$. \formalized}
    \end{itemize}
  \end{changemargin}
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{A modal language for harnessing the multiverse}
  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \vspace*{-0.3em}
  \begin{itemize}
    \small
    \item \hil{everywhere} \tabto{2.05cm}($\necessary\varphi$)\tabto{2.98cm}
    iff it holds \hil{in every topos}
    (over the current base).
    \item \hil{somewhere} \tabto{2.05cm}($\possible\varphi$)\tabto{2.98cm} iff
    it holds \hil{in some positive topos}.
    \item \hil{proximally} \tabto{2.05cm}($\xpossible\varphi$)\tabto{2.98cm}
    iff it holds \hil{in some positive overt topos}.
  \end{itemize}

  We then have:
  \begin{enumerate}
    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
    \necessary\mathsf{Well}\sinf(X,{\le})$. \formalized
    \item $(\possible\varphi) \Longleftrightarrow \varphi$,
    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
    in~$\%$).~\formalized
    \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
    if~$\varphi$ is a bounded first-order statement. \formalized
    \item For every inhabited type~$X$, $\xpossible\necessary
    \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
    \to X\_ \mathsf{Surjective}(f))$. \formalized

@@

\begin{frame}{The plain combinatorics of toposes}
  \begin{enumerate}\justifying
    \item Realize a generic gadget as some kind of limit of \hil{approximations}
    from the base universe. For the generic function $f_0 : \NN \to X$:

@@

    \pause
    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
    gadget as assertions about its approximations~(\formalized).

@@

    \pause
    \item Be prepared to \hil{evolve approximations} (\formalized).

@@

\begin{frame}{Formal metatheory}
  \small
  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
  \begin{itemize}
    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
    arbitrary small categories, and their corresponding sheaf models

@@

{\renewcommand{\insertframeextra}{a}
\begin{frame}{Increasing subsequences as convenient fictions}
  \justifying
  Let~$X$ be a quasi-order.
  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.

@@

  \begin{block}{Constructive reimagination}
    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
    then~$\nabla\,B\,[\,]$. \formalized
  \end{block}
  \vspace*{-0.4em}
  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  By assumption, this subsequence has a finite prefix validating~$B$. This
  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
  has a prefix validating~$B$ by monotonicity \emph{there}. So
  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
  actually is a finite prefix validating~$B$. \qed\par}
\end{frame}}

>>>>

% compile using: agda --latex; result will then be in the latex/ subfolder
\documentclass[11pt,utf8,notheorems,compress,t,aspectratio=169]{beamer}
\usepackage{etex}

@@

% Workaround for the issue described at
% https://tex.stackexchange.com/questions/164406/beamer-using-href-in-notes.
\newcommand{\fixedhref}[2]{\parbox[t][0pt][t]{0pt}{\vspace*{-\paperheight}\vspace*{-0.6em}\href{#1}{#2}}\href{#1}{#2}}

@@

\title{Towards topological type theory for decrypting transfinite methods in classical mathematics}

\author{Ingo Blechschmidt}
\date{June 9th, 2025}

\setbeameroption{show notes on second screen=bottom}
\newcommand{\jnote}[2]{\only<#1>{\note{\setlength\parskip{\medskipamount}\footnotesize\justifying#2\par}}}
\makeatletter
\setbeamertemplate{note page}
{%
  {%
    \scriptsize
    \insertvrule{.25\paperheight}{white!90!black}
    \vskip-.25\paperheight
    \nointerlineskip
    \vbox{
      %\hfill\insertslideintonotes{0.25}\hskip-\Gm@rmargin\hskip0pt%
      %\vskip-0.25\paperheight%
      %\nointerlineskip
      \begin{pgfpicture}{0cm}{0cm}{0cm}{0cm}
        \begin{pgflowlevelscope}{\pgftransformrotate{90}}
          {\pgftransformshift{\pgfpoint{-2cm}{0.2cm}}%
          \pgftext[base,left]{\footnotesize\the\year-\ifnum\month<10\relax0\fi\the\month-\ifnum\day<10\relax0\fi\the\day}}
        \end{pgflowlevelscope}
      \end{pgfpicture}}
    \nointerlineskip
    \vbox to .25\paperheight{\vskip0.5em
      \hbox{\insertshorttitle[width=8cm]}%
      \setbox\beamer@tempbox=\hbox{\insertsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertsubsection}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip17.4pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\begin{minipage}[t]{7.5cm}\def\breakhere{}\insertsubsection\end{minipage}}\fi%
      }%
      \setbox\beamer@tempbox=\hbox{\insertshortframetitle}%
      \hbox{\ifdim\wd\beamer@tempbox>1pt{\hskip30.8pt\raise3pt\hbox{\vrule
            width0.4pt height7pt\vrule width 9pt
            height0.4pt}}\hskip1pt\hbox{\insertshortframetitle[width=7cm]}\fi%
      }%
      \vfil}%
  }%
  \vskip.25em
  \nointerlineskip
  \insertnote
}
\makeatother

@@

\newunicodechar{∇}{\ensuremath{\nabla}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{σ}{\ensuremath{\sigma}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{∷}{\ensuremath{\dblcolon}}
\newunicodechar{⧺}{\ensuremath{\!\!\!}}
\newunicodechar{₁}{\ensuremath{_1}}
\newunicodechar{₂}{\ensuremath{_2}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{ℕ}{\ensuremath{\NN}}
\newunicodechar{↠}{\ensuremath{\twoheadrightarrow}}
\newunicodechar{ʳ}{\ensuremath{^r}}

@@

\addtocounter{framenumber}{-1}
\begin{frame}\frametitle{A classical logic fairy tale}
  \centering\bigskip\bigskip\bigskip
  \includegraphics[height=9.5em]{philosophers-stone}

  \jnote{1}{
    \vspace*{-1em}
    \begin{changemargin}{-1.5em}{-1.5em}
      \setlength{\columnsep}{15pt}
      \begin{multicols}{2}
        \justifying
        \scriptsize

        \textbf{Narrator.} Once upon a time, in a kingdom far, far away, the
        queen of the country and of all Möbius strips called for her royal
        philosopher.

        \textbf{Queen.} Philosopher! I ask you to carry out the following order.
        Get me the Philosopher's Stone, or alternatively find out how one could
        produce arbitrary amounts of gold with it!

        \textbf{Philosopher.} But my queen! I haven't studied anything useful!
        How could I fulfill this order?

        \textbf{Queen.} That is not my concern. I'll see you again tomorrow.
        Should you not accomplish the task, I will take your head off.

        \textbf{Narrator.} After a long and wakeful night the philosopher was
        called to the queen again.

        \textbf{Queen.} Tell me! What do you have to report?

        \textbf{Philosopher.} It was not easy and I needed to follow lots of
        obscure references, but finally I actually found out how to use the
        Philosopher's Stone to produce arbitrary amounts of gold. But only I can
        conduct this procedure, your royal highness.

        \textbf{Queen.}
        Alright. So be it.

        \textbf{Narrator.} And so years passed by, during which the philosopher
        imagined herself to be safe. The queen searched for the stone on her own,
        but as long as she hadn't found it, the philosopher didn't need to worry.
        Yet one day the impossible happened: The queen has found the stone! And
        promptly called for her philosopher.

        \textbf{Queen.} Philosopher, look! I have found the Philosopher's Stone!
        Now live up to your promise! \emph{[She hands over the stone.]}

        \textbf{Philosopher.} Thank you. \emph{[She inspects the stone.]} This is
        indeed the Philosopher's Stone. Many years ago you asked me to either
        acquire the Philosopher's Stone or find out how to produce arbitrary
        amounts of gold using it. Now it's my pleasure to present to you the
        Philosopher's Stone. \emph{[She returns the stone.]}
      \end{multicols}
    \end{changemargin}
  }

  \jnote{2}{
    Adapted from
    \fixedhref{http://blog.ezyang.com/2013/04/a-classical-logic-fairy-tale/}{Edward
    Yang's blog}.

    The idea conveyed in the fairy tale is at the core of the
    \emph{constructive} proof of $\neg\neg(A \vee \neg A)$. While the law of
    excluded middle (\textsc{lem}) is not available in constructive
    mathematics, up to a double negation every instance of it is.

    This observation gives rise to the double negation translation, a
    transformation of (statements and) proofs in classical logic to proofs in
    intuitionistic logic, at the only expense of introducing double negations;
    by following this up with Friedman's trick with the nontrivial exit
    continuation (``Baby Barr Theorem''), we can in many cases even get rid of
    these double negations in a second step.

    Getting rid of applications of the law of excluded middle is in many
    contexts (an amazing mind-boggingly wonder at first, but then) a routine
    matter---see \href{https://rt.quasicoherent.io/}{here} for more details and
    pointers to the literature.
  }
\end{frame}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{1.7em}\includegraphics[height=\paperheight]{sea-of-clouds-1}\end{minipage}}
\begin{frame}{A case study in Hilbert's program}
  \jnote{1}{
    Well quasi-orders are an important notion in proof theory and termination
    analysis.
  }

  \jnote{2-3}{
    \vspace*{-0.5em}
    The presented proof rests on the law of excluded middle and hence cannot
    immediately be interpreted as a program for finding suitable indices~$i <
    j$. However, constructive proofs are also possible (for instance by
    induction on the value of a given term of the sequence, see
    \href{https://www.math.lmu.de/~petrakis/Dickson.pdf}{Constructive
    combinatorics of Dickson's Lemma} by Iosif Petrakis for several fine
    quantitative results). And even more:
    There is a procedure for regarding this proof---and many others in the
    theory of well quasi-orders---as \emph{blueprints} for more informative
    constructive proofs. This shall be our motto for today:

    \emph{Do not take classical proofs literally, instead ask which
    constructive proofs they are blueprints for.}
  }

  \jnote{3}{
    The displayed stability results, along with several others, provide a
    flexible toolbox for constructing new well quasi-orders from given ones.
    However, with the classical formulation of \emph{well},
    shortly to be renamed~``well\sinf'', these results are
    \href{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Orthogonal.html}{inherently
    classical} (even the weaker claim ``the product of two streamless sets is
    streamless'' admits countermodels).

    (In Higman's lemma, the set~$X^*$ of finite lists of elements of~$X$ is
    equipped with the following ordering: We have~$(x_0 \dblcolon \ldots
    \dblcolon x_{n-1} \dblcolon [\,]) \leq (y_0 \dblcolon \ldots \dblcolon
    y_{m-1} \dblcolon [\,])$ iff there is an increasing injection~$f : \{
    0,\ldots,n-1 \} \to \{ 0,\ldots,m-1 \}$ such that~$x_i \leq y_{f(i)}$ for all~$i < n$.
    Kruskal's theorem is the result behind the celebrated enormous number
    $\mathsf{TREE}(3)$.)
  }

  \jnote{7}{
    Thanks to work by Thierry Coquand, Daniel Fridlender and Monika
    Seisenberger, a constructive substitute is available, the notion well\ind.
    In classical mathematics (where bar induction is available), this notion is
    equivalent to well\sinf.

    The predicate~$\mathsf{Good}$ of finite lists~$\sigma = (x_0 \dblcolon
    \ldots \dblcolon x_{n-1} \dblcolon [\,])$ appearing in the definition
    of~$G$ expresses that some earlier term is at most some later term,
    i.\,e.\@ that there merely exist~$i < j$ such that~$x_i \leq x_j$.

    The operator $\dblcolon^r$ is adding an element at the end.
  }

  \jnote{8}{
    \begin{columns}[t]
      \begin{column}{0.65\textwidth}
        The original notion well\sinf:
        \begin{itemize}
          \justifying
          \item[\cmark] short and simple
          \\[-1.3em]
          \item[\cmark] constructively satisfied for the main examples (but
          only because of the theory around well\ind)
          \\[-1.3em]
          \item[\cmark] concise abstract proofs (albeit employing transfinite methods)
          \\[-1.3em]
          \item[\xmark] main results not constructively attainable
          \\[-1.3em]
          \item[\xmark] philosophically strenuous by the quantification over all sequences
          \\[-1.3em]
          \item[\xmark] not stable under ``change of base''---a forcing extension
          of the universe may well contain more sequences than the base universe
          \\[-1.3em]
          \item[\xmark] negative (universal) condition
        \end{itemize}
      \end{column}

      \begin{column}{0.33\textwidth}
        The constructive substitute well\ind:
        \begin{itemize}
          \justifying
          \item[\cmark] main results constructive
          \\[-1.3em]
          \item[\cmark] stable under change of base
          \\[-1.3em]
          \item[\cmark] positive (existential) condition
          \\[-1.3em]
          \item[\xmark] proofs intriguing, but also somewhat alien, not just some
          trivial reshuffling of the classical arguments; classical sequence
          language cannot be used
        \end{itemize}
      \end{column}
    \end{columns}
  }

  \textbf{Def.} Let~$(X,{\leq})$ be a quasi-order.
  \vspace*{-0.3em}
  \begin{itemize}
    \item A sequence~$\alpha : \NN \to X$ is \hil{good} iff
    there merely exist~$i < j$ with~$\alpha\,i \leq \alpha\,j$.
    \item The quasi-order~$X$ is \hil{well\only<4->{\sinf}} iff every sequence~$\NN \to X$ is good.
  \end{itemize}
  \pause
  \vspace*{-1.3em}

@@

  \textbf{Def.} A quasi-order~$X$ is \hil{well\ind} iff~$G\,[\,]$, where $G$ is
  the following inductively defined predicate on \hil{finite lists}.~\formalized{}
  (In presence of \bad{bar induction}, $\text{well}\ind \Leftrightarrow
  \text{well}\sinf$.)
  \[
    \small
    \infer{\mathsf{now}\,p : G\,\sigma}{p : \mathsf{Good}\,\sigma}
    \qquad
    \infer{\mathsf{later}\,f : G\,\sigma}{f : (x : X) \to G\,(\sigma \dblcolon^r x)}
  \]
  \vspace*{-1.3em}

  \only<8>{
    \begin{columns}[c]
      \begin{column}{0.01\textwidth}
	\includegraphics[height=2.4em]{question-mark}
      \end{column}
      \begin{column}{0.9\textwidth}
      Is there a procedure for reinterpreting \hil{classical proofs}
      regarding well\sinf{} as \newline \hil{blueprints for constructive proofs} regarding well\ind?
      \end{column}
    \end{columns}
  }
\end{frame}}

\begin{frame}{Missing functions in the type of all functions?}
  \jnote{1}{
    The double negation translation technique fails to constructivize the two
    proofs presented on the slide.

    The reason is that, after applying the translation, the ``$I$-extracted
    subsequence'' appearing in the proof on the left will no longer actually be
    a function~$\NN \to X$. Instead, it will be a partially defined function
    with the property that each number~$n$ does \emph{not~not} belong to the
    domain. However, the assumption that~$X$ is well\sinf{} does not apply to
    these kinds of generalized sequences.

    In contrast, the notion well\ind{} is much stronger. If~$X$ is well\ind,
    then not only will every actual sequence~$\NN \to X$ be good, but (in a
    suitable sense) so will every up-to-$\neg\neg$-partially-defined sequence,
    every multivalued sequence and many more kinds of generalized sequences.
  }

  \begin{columns}[t]
    \begin{column}{0.57\textwidth}
      \begin{block}{Behold: A transfinite tool \ldots\phantom{g}}
        \justifying
        \textbf{Lemma.} \badbox{\textsc{lem}} Let~$X$ be well\sinf.
        Let~$\alpha : \NN \to X$. Then there is an increasing
        subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
      \end{block}

@@

    In set and type theory, we can also enlarge \hil{the mathematical universe}:
    \begin{itemize}
      \small
      \item Force a \hil{generic sequence} $\NN \to X$.
            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Cohen.html}{\formalized}
      \item Force a \hil{generic enumeration} $\NN \twoheadrightarrow X$
            (even if $X$ is uncountable).
            \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}
      \item Force a \hil{generic prime ideal} of a given ring.
            \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html}{\formalized}
    \end{itemize}
    \pause

    Central observations of the multiversal yoga:
    \begin{itemize}
      \small
      \item A quasi-order~$X$ is well\ind iff the generic sequence~$\NN \to X$
      is good.
      \item A set is Noetherian iff the generic sequence~$\NN \to X$
      has two repeated terms.
      \item A relation is well-founded iff for the generic descending chain,
      $\bot$.
      \item A ring element is nilpotent iff it is contained in the generic
      prime ideal.
      \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Prime.html\#corollary}{\formalized}
    \end{itemize}
    \pause

    The mystery of nongeometric sequents:
    \begin{itemize}
    \small
      \item The generic ring is a field.
      \item \ \\[-1.2em]\mbox{For the generic surjection $f : \NN \twoheadrightarrow \RR$,
      $\neg\neg \exists n \? \NN\_ f(n) = \pi \wedge f(n+1) = e$.
      \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}}
    \end{itemize}
  \end{changemargin}

  \jnote{1}{
    The technique for enlarging the mathematical universe is called
    \emph{forcing} and was originally pioneered by Paul Cohen. Set theorists
    use forcing to construct new models of set theory from given ones, in order
    to explore the range of set-theoretic possibility. For instance, by forcing
    we can construct models of~\textsc{zfc} validating the continuum hypothesis
    and also models which falsify it.

    We here refer to a simplification of original forcing which is useful in a
    constructive metatheory. At its core, every forcing extension is just a
    formula and proof translation of a certain form. For instance, there is a
    forcing extension validating~\bad{\textsc{lem}} even if the base universe
    does not; this forcing extension is not a deep mystery, for a statement holds
    in that forcing extension iff its double negation translation holds in the
    base universe and it is well-known that the double negation translation
    of every instance of~\bad{\textsc{lem}} is an intuitionistic tautology.

    \href{https://www.speicherleck.de/iblech/stuff/slides-herrsching2023.pdf}{Here
    is a set of slides on constructive forcing}, and Section~4 of
    \href{https://raw.githubusercontent.com/iblech/constructive-maximal-ideals/master/tex/extended.pdf}{this
    joint paper with Peter Schuster} contains a written summary of constructive
    forcing.
  }

  \jnote{2}{
    Forcing the existence of a new function~$\NN \to X$ is also called
    \emph{Cohen forcing}. Forcing the existence of a new surjection is called
    \emph{Lévy collapse}. Forcing the existence of a prime ideal does not have
    an established name, perhaps it should be called \emph{Zariski forcing}; it
    has been explored by Thierry Coquand and others of the constructive algebra
    crew.

    The multiversal yoga allows us to reason with the good notion well\ind{}
    using the simple language of well\sinf, and similarly with several other
    notions.
  }

  \jnote{3}{
    For more on the mystery of nongeometric sequents, see
    \href{https://www.speicherleck.de/iblech/stuff/slides-birmingham2019.pdf}{this
    set of slides}.
  }
\end{frame}}

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{4.95cm}\includegraphics[width=\paperwidth]{topos-horses-lighter}\end{minipage}}
\begin{frame}{A modal language for harnessing the multiverse}
  \jnote{1}{
    By \emph{topos}, we mean \emph{Grothendieck topos}. In constructive
    forcing, a ``forcing extension of the base universe'' is exactly the same
    thing as a Grothendieck topos.

    A particular member of the rich and varied landscape of toposes is the
    \emph{trivial topos}, in which every statement whatsoever holds. By
    restricting to positive toposes, we exclude this special case.

    For positive toposes~$\E$, a geometric implication holds in~$\E$ iff it
    holds in the base universe. For positive overt toposes~$\E$, we even have
    that a bounded first-order formula holds in~$\E$ iff it holds in the base.
    Hence, for the purpose of verifying a bounded first-order assertion about
    the base, we can freely pass to a positive overt topos with problem-adapted
    better higher-order properties (such as that some uncountable set from the
    base now appears countable, or that an infinite sequence whose existence is
    predicted by failing dependent choice now actually exists).

    \href{https://www.speicherleck.de/iblech/stuff/early-draft-modal-multiverse.pdf}{Here
    is a rough early draft of a preprint with more details about the modal
    multiverse.}
  }

  \jnote{2}{
    The idea to study the modal multiverse of toposes in a principled manner
    was proposed by Alexander Oldenziel in 2016. \emph{Foreshadowed by:}
    \begin{itemize}
      \item[1984\phantom{s}] André Joyal, Miles Tierney. ``An extension of the Galois theory of Grothendieck''.
      \\[-1em]
      \item[1987\phantom{s}] Andreas Blass. ``Well-ordering and induction in intuitionistic logic and topoi''.
      \\[-1em]
      \item[2010s] Milly Maietti, Steve Vickers. Ongoing work on arithmetic universes.
      \\[-1em]
      \item[2011\phantom{s}] Joel David Hamkins. ``The set-theoretic multiverse''.
      \\[-1em]
      \item[2013\phantom{s}] Shawn Henry. ``Classifying topoi and preservation of higher order logic by geometric morphisms''.
    \end{itemize}
  }

  \jnote{3}{
    With the modal language we seek to provide an accessible and modular
    framework for constructivization results.

    For instance, conservativity of classical logic over intuitionistic logic
    for geometric implications (known under various names such as Barr's
    theorem, Friedman's trick, escaping the continuation monad, \ldots) is
    packaged up by the observation that \emph{somewhere}, the law of excluded
    middle holds.

    Another example:
    In the community around Krull's lemma, it is well-known that
    we can constructively infer that a given ring element~$x \in A$ is nilpotent from
    knowing that it is contained in the \emph{generic prime ideal} of~$A$. This entity
    is not actually an honest prime ideal of the ring~$A$ in the base
    universe, but a certain combinatorial notion (efficiently dealt with using
    \emph{entailment relations}). Constructive forcing allows us to reify the
    generic prime ideal as an actual prime ideal in a suitable forcing extension,
    so in a suitable topos (a version of the little Zariski topos of the ring).
  }

  \textbf{Def.} A statement~$\varphi$ holds \ldots
  \vspace*{-0.3em}
  \begin{itemize}
    \small
    \item \hil{everywhere} \tabto{2.05cm}($\necessary\varphi$)\tabto{2.98cm}
    iff it holds \hil{in every topos}
    (over the current base).
    \item \hil{somewhere} \tabto{2.05cm}($\possible\varphi$)\tabto{2.98cm} iff
    it holds \hil{in some positive topos}.
    \item \hil{proximally} \tabto{2.05cm}($\xpossible\varphi$)\tabto{2.98cm}
    iff it holds \hil{in some positive overt topos}.
  \end{itemize}

  We then have:
  \begin{enumerate}
    \item $\mathsf{Well}\ind(X,{\le}) \Longleftrightarrow
    \necessary\mathsf{Well}\sinf(X,{\le})$.
    \item $(\possible\varphi) \Longleftrightarrow \varphi$,
    if $\varphi$ is a geometric implication (``$\forall\ldots\forall (\%
    \Rightarrow \%)$'', with no~$\forall$ nor~$\Rightarrow$
    in~$\%$).~\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html\#CoherentEquivalence}{\formalized}
    \item $(\xpossible\varphi) \Longleftrightarrow \varphi$,
    if~$\varphi$ is a bounded first-order statement.
    \fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html\#1\%CB\%A2\%E1\%B5\%97OrderEquivalence}{\formalized}
    \item For every inhabited type~$X$, $\xpossible\necessary
    \mathsf{Countable}(X)$, where~$\mathsf{Countable}(X) \defeqv (\exists f : \NN
    \to X\_ \mathsf{Surjective}(f))$.
    \fixedhref{https://lets-play-agda.quasicoherent.io/Padova2025.Explorations.Forcing.Levy.html}{\formalized}

@@

\begin{frame}{The plain combinatorics of toposes}
  \jnote{3-}{
    Proofs (about the base universe) are (pure) programs; proofs about a forcing
    extension are effectful programs unfolding in the~$\nabla$ monad.

    The $\nabla$ monad is a refinement of the (monotonic) state monad.
  }

  \begin{enumerate}\justifying
    \item Realize a generic gadget as some kind of limit of \hil{approximations}
    from the base universe. For the generic function $f_0 : \NN \to X$:

@@

    \pause
    \item \hil{Reinterpret}, in a mechanical fashion, assertions about the generic
    gadget as assertions about its
    approximations~(\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Semantics.html}{\formalized}).

@@

    \pause
    \item Be prepared to \hil{evolve approximations}
    (\fixedhref{https://iblech.github.io/constructive-maximal-ideals/Forcing.Base.html}{\formalized}).

@@

\begin{frame}{Formal metatheory}
  \jnote{2}{
    \href{https://msp.cis.strath.ac.uk/types2025/slides/TYPES2025-slides57.pdf}{Exciting
    ongoing work on split type theory} by Martin Baillon, Assia Mahboubi and
    Pierre-Marie Pédrot!
  }

  \small
  \good{\cmark{}} \emph{There are type-theoretic multiverses,} such as
  \begin{itemize}
    \item the collection of all $\mathrm{PSh}(\mathcal{C} \times \mathcal{B})$,
    where~$\mathcal{B}$ ranges over cube categories and~$\mathcal{C}$ over
    arbitrary small categories, and their corresponding sheaf models

@@

{\renewcommand{\insertframeextra}{a}
\begin{frame}{Increasing subsequences as convenient fictions}
  \jnote{1}{
    On this slide~$\nabla$ is the monad described on slide 5,
    and~$\nabla^\nnearrow$ is the variant where lists are only allowed to grow
    in a monotonically increasing manner; so~$\nabla$ is the monad for the
    generic sequence~$\NN \to X$, while~$\nabla^\nnearrow$ is the monad for the
    generic increasing sequence~$\NN \to X$.

    The displayed multiversal proof gives a positive answer to a question posed
    by Stefano Berardi, Gabriele Buriola and Peter Schuster, see
    \href{https://www.speicherleck.de/iblech/stuff/slides-abmv2024.pdf\#page=16}{this
    set of slides}.

    At the end of the day, this proof yields a concrete algorithm (see
    \href{https://www.speicherleck.de/iblech/stuff/dickson/implicational.pl}{here}
    for manually-extracted Perl code). Given an infinite sequence, this
    algorithm seeks, in a certain systematic but not exhaustive manner, a
    sufficiently long monotonic subsequence, making judicious use of
    backtracking. This search terminates, though this fact is not obvious just
    by looking at the algorithm.
  }

  \justifying
  Let~$X$ be a quasi-order.
  Let~$B : \mathsf{List}\,X \to \mathsf{Prop}$ be a monotone predicate.

@@

  \begin{block}{Constructive reimagination}
    \textbf{Thm.} If~$X$ is well\ind and if~$\nabla^\nnearrow\,B\,[\,]$,
    then~$\nabla\,B\,[\,]$.
    \fixedhref{https://www.speicherleck.de/iblech/stuff/dickson//Dickson.html}{\formalized}
  \end{block}
  \vspace*{-0.4em}
  {\small\emph{Proof.} Let~$\alpha : \NN \to X$ be a sequence (in an arbitrary topos).
  \emph{Somewhere,} \textsc{lem} holds. \emph{There}~$X$
  is still well\sinf, so that we have an increasing subsequence~$\alpha\,i_0 \leq \alpha\,i_1 \leq \cdots$.
  By assumption, this subsequence has a finite prefix validating~$B$. This
  prefix is part of a prefix of the original sequence~$\alpha$. Hence~$\alpha$
  has a prefix validating~$B$ by monotonicity \emph{there}. So
  \emph{somewhere} there is a finite prefix validating~$B$. Thus there
  actually is a finite prefix validating~$B$. \qed\par}
\end{frame}}
